---
pagetitle: Notebook report
output:
  html_notebook:
    highlight: zenburn
    theme: cosmo
    toc: yes
    code_folding: hide
---

```{r r_setup, include = FALSE}
## initial settings
knitr::opts_chunk$set(
  comment = NA,
  echo = TRUE,
  error = TRUE,
  cache = FALSE,
  message = FALSE,

  dpi = 96,
  warning = FALSE
)

## width to use when printing tables etc.
options(
  width = 250,
  scipen = 100,
  max.print = 5000,
  stringsAsFactors = FALSE
)

## make all required libraries available by loading radiant package if needed
if (is.null(shiny::getDefaultReactiveDomain())) library(radiant)

## include code to load the data you require
## for interactive use attach the r_data environment
# attach(r_data)
```

<style>
.btn, .form-control, pre, code, pre code {
  border-radius: 4px;
}
.table {
  width: auto;
}
ul, ol {
  padding-left: 18px;
}
code, pre, pre code {
  overflow: auto;
  white-space: pre;
  word-wrap: normal;
}
code {
  color: #c7254e;
  background-color: #f9f2f4;
}
pre {
  background-color: #ffffff;
}
</style>

## Marketing Analytics HW#1

#### Question #1 - T-Test - "dormant" 


```{r fig.width = 7, fig.height = 4.31, dpi = 96}
result <- compare_means(
  karlan_list_2007_rl, 
  var1 = "treatment_factor", 
  var2 = "dormant"
)
summary(result, show = TRUE)
plot(result, plots = "scatter", custom = FALSE)
```

#### Question #1 - Linear Regression - "dormant"



```{r fig.width = 7, fig.height = 2.69, dpi = 96}
result <- regress(
  karlan_list_2007_rl, 
  rvar = "dormant", 
  evar = "treatment_factor"
)
summary(result)
plot(result, plots = "scatter", custom = FALSE)
```


#### Question #1 - T-Test - "pblack"


```{r fig.width = 7, fig.height = 4.31, dpi = 96}
result <- compare_means(
  karlan_list_2007_rl, 
  var1 = "treatment_factor", 
  var2 = "pblack"
)
summary(result, show = TRUE)
plot(result, plots = "scatter", custom = FALSE)
```
#### Question #1 - Linear Regression - "pblack"

```{r fig.width = 7, fig.height = 2.69, dpi = 96}
result <- regress(
  karlan_list_2007_rl, 
  rvar = "pblack", 
  evar = "treatment_factor"
)
summary(result)
plot(result, plots = "scatter", nrobs = -1, custom = FALSE)
```



#### Question #1 - T-Test - "median_hhincome"


```{r fig.width = 7, fig.height = 4.31, dpi = 96}
result <- compare_means(
  karlan_list_2007_rl, 
  var1 = "treatment_factor", 
  var2 = "median_hhincome"
)
summary(result, show = TRUE)
plot(result, plots = "scatter", custom = FALSE)

```

#### Question #1 - Linear Regression - "median_hhincome"



```{r fig.width = 7, fig.height = 2.69, dpi = 96}
result <- regress(
  karlan_list_2007_rl, 
  rvar = "median_hhincome", 
  evar = "treatment_factor"
)
summary(result)
plot(result, plots = "scatter", nrobs = -1, custom = FALSE)
```

#### Question #2 - Proportion of People Who Donated by Treatment vs Control Group


```{r fig.width = 7, fig.height = 7, dpi = 96}
visualize(
  karlan_list_2007_rl, 
  xvar = "treatment_factor", 
  yvar = "gave", 
  type = "bar", 
  axes = "flip", 
  labs = list(
    title = "Proportion of People Who Donated by Treatment vs Control Group", 
    x = "Treatment Group", 
    y = "% Who Donated"
  ), 
  custom = FALSE
)
```
#### Question #2 - T-Test - Gave by Treatment vs Control Group


```{r fig.width = 7, fig.height = 4.31, dpi = 96}
result <- compare_means(
  karlan_list_2007_rl, 
  var1 = "treatment_factor", 
  var2 = "gave"
)
summary(result, show = TRUE)
plot(result, plots = "bar", custom = FALSE)

```

#### Question #2 - Regression - Gave by Treatment vs Control Group




```{r fig.width = 7, fig.height = 2.69, dpi = 96}
result <- regress(
  karlan_list_2007_rl, 
  rvar = "gave", 
  evar = "treatment_factor"
)
summary(result)
plot(result, plots = "pred_plot", incl = "treatment_factor", custom = FALSE)
```

