---
title: "Multinomial Logit Model"
author: "Rob Loftus"
date: May 27, 2025

---



This assignment explores two methods for estimating the MNL model: (1) via Maximum Likelihood, and (2) via a Bayesian approach using a Metropolis-Hastings MCMC algorithm. 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyverse)
library(mlogit)
df <- read_csv("conjoint_data.csv")
```

## 1. Likelihood for the Multi-nomial Logit (MNL) Model

Suppose we have $i=1,\ldots,n$ consumers who each select exactly one product $j$ from a set of $J$ products. The outcome variable is the identity of the product chosen $y_i \in \{1, \ldots, J\}$ or equivalently a vector of $J-1$ zeros and $1$ one, where the $1$ indicates the selected product. For example, if the third product was chosen out of 3 products, then either $y=3$ or $y=(0,0,1)$ depending on how we want to represent it. Suppose also that we have a vector of data on each product $x_j$ (eg, brand, price, etc.). 

We model the consumer's decision as the selection of the product that provides the most utility, and we'll specify the utility function as a linear function of the product characteristics:

$$ U_{ij} = x_j'\beta + \epsilon_{ij} $$

where $\epsilon_{ij}$ is an i.i.d. extreme value error term. 

The choice of the i.i.d. extreme value error term leads to a closed-form expression for the probability that consumer $i$ chooses product $j$:

$$ \mathbb{P}_i(j) = \frac{e^{x_j'\beta}}{\sum_{k=1}^Je^{x_k'\beta}} $$

For example, if there are 3 products, the probability that consumer $i$ chooses product 3 is:

$$ \mathbb{P}_i(3) = \frac{e^{x_3'\beta}}{e^{x_1'\beta} + e^{x_2'\beta} + e^{x_3'\beta}} $$

A clever way to write the individual likelihood function for consumer $i$ is the product of the $J$ probabilities, each raised to the power of an indicator variable ($\delta_{ij}$) that indicates the chosen product:

$$ L_i(\beta) = \prod_{j=1}^J \mathbb{P}_i(j)^{\delta_{ij}} = \mathbb{P}_i(1)^{\delta_{i1}} \times \ldots \times \mathbb{P}_i(J)^{\delta_{iJ}}$$

Notice that if the consumer selected product $j=3$, then $\delta_{i3}=1$ while $\delta_{i1}=\delta_{i2}=0$ and the likelihood is:

$$ L_i(\beta) = \mathbb{P}_i(1)^0 \times \mathbb{P}_i(2)^0 \times \mathbb{P}_i(3)^1 = \mathbb{P}_i(3) = \frac{e^{x_3'\beta}}{\sum_{k=1}^3e^{x_k'\beta}} $$

The joint likelihood (across all consumers) is the product of the $n$ individual likelihoods:

$$ L_n(\beta) = \prod_{i=1}^n L_i(\beta) = \prod_{i=1}^n \prod_{j=1}^J \mathbb{P}_i(j)^{\delta_{ij}} $$

And the joint log-likelihood function is:

$$ \ell_n(\beta) = \sum_{i=1}^n \sum_{j=1}^J \delta_{ij} \log(\mathbb{P}_i(j)) $$



## 2. Simulate Conjoint Data

We will simulate data from a conjoint experiment about video content streaming services. We elect to simulate 100 respondents, each completing 10 choice tasks, where they choose from three alternatives per task. For simplicity, there is not a "no choice" option; each simulated respondent must select one of the 3 alternatives. 

Each alternative is a hypothetical streaming offer consistent of three attributes: (1) brand is either Netflix, Amazon Prime, or Hulu; (2) ads can either be part of the experience, or it can be ad-free, and (3) price per month ranges from \$4 to \$32 in increments of \$4.

The part-worths (ie, preference weights or beta parameters) for the attribute levels will be 1.0 for Netflix, 0.5 for Amazon Prime (with 0 for Hulu as the reference brand); -0.8 for included adverstisements (0 for ad-free); and -0.1*price so that utility to consumer $i$ for hypothethical streaming service $j$ is 

$$
u_{ij} = (1 \times Netflix_j) + (0.5 \times Prime_j) + (-0.8*Ads_j) - 0.1\times Price_j + \varepsilon_{ij}
$$

where the variables are binary indicators and $\varepsilon$ is Type 1 Extreme Value (ie, Gumble) distributed.

The following code provides the simulation of the conjoint data.

:::: {.callout-note collapse="true"}
```{r}

# set seed for reproducibility
set.seed(123)

# define attributes
brand <- c("N", "P", "H") # Netflix, Prime, Hulu
ad <- c("Yes", "No")
price <- seq(8, 32, by=4)

# generate all possible profiles
profiles <- expand.grid(
    brand = brand,
    ad = ad,
    price = price
)
m <- nrow(profiles)

# assign part-worth utilities (true parameters)
b_util <- c(N = 1.0, P = 0.5, H = 0)
a_util <- c(Yes = -0.8, No = 0.0)
p_util <- function(p) -0.1 * p

# number of respondents, choice tasks, and alternatives per task
n_peeps <- 100
n_tasks <- 10
n_alts <- 3

# function to simulate one respondent’s data
sim_one <- function(id) {
  
    datlist <- list()
    
    # loop over choice tasks
    for (t in 1:n_tasks) {
        
        # randomly sample 3 alts (better practice would be to use a design)
        dat <- cbind(resp=id, task=t, profiles[sample(m, size=n_alts), ])
        
        # compute deterministic portion of utility
        dat$v <- b_util[dat$brand] + a_util[dat$ad] + p_util(dat$price) |> round(10)
        
        # add Gumbel noise (Type I extreme value)
        dat$e <- -log(-log(runif(n_alts)))
        dat$u <- dat$v + dat$e
        
        # identify chosen alternative
        dat$choice <- as.integer(dat$u == max(dat$u))
        
        # store task
        datlist[[t]] <- dat
    }
    
    # combine all tasks for one respondent
    do.call(rbind, datlist)
}

# simulate data for all respondents
conjoint_data <- do.call(rbind, lapply(1:n_peeps, sim_one))

# remove values unobservable to the researcher
conjoint_data <- conjoint_data[ , c("resp", "task", "brand", "ad", "price", "choice")]

# clean up
rm(list=setdiff(ls(), "conjoint_data"))
```
::::


## 3. Preparing the Data for Estimation

The "hard part" of the MNL likelihood function is organizing the data, as we need to keep track of 3 dimensions (consumer $i$, covariate $k$, and product $j$) instead of the typical 2 dimensions for cross-sectional regression models (consumer $i$ and covariate $k$). The fact that each task for each respondent has the same number of alternatives (3) helps.  In addition, we need to convert the categorical variables for brand and ads into binary variables.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

# _todo: reshape and prep the data_

# Step 1: Prepare data
conjoint_data <- conjoint_data %>%
  mutate(
    ad = ifelse(ad == "Yes", 1, 0),
    chid = paste(resp, task, sep = "_")
  )

# Step 2: Create alt ID
conjoint_data <- conjoint_data %>%
  group_by(chid) %>%
  mutate(alt_id = row_number()) %>%
  ungroup()

# Step 3: Convert brand to factor - Hulu baseline
conjoint_data <- conjoint_data %>%
  mutate(brand = factor(brand, levels = c("H", "P", "N")))

# Step 4: Use mlogit.data with alt ID
mlogit_df <- mlogit.data(
  data = conjoint_data,
  choice = "choice",
  shape = "long",
  chid.var = "chid",
  alt.var = "alt_id", 
  id.var = "resp"
)


```

## 4. Estimation via Maximum Likelihood

Here are the results from our first model, which used a standard statistical method called Maximum Likelihood Estimation (MLE). These estimates tell us how each feature (brand, ads, price) affects people’s choices.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

# _todo: Code up the log-likelihood function._

# Filter and arrange data
library(dplyr)

X_data <- conjoint_data %>%
  mutate(
    netflix = ifelse(brand == "N", 1, 0),
    prime = ifelse(brand == "P", 1, 0)
  ) %>%
  arrange(resp, task, alt_id) 

# Step 2

log_likelihood <- function(params, data) {
  beta_netflix <- params[1]
  beta_prime   <- params[2]
  beta_ads     <- params[3]
  beta_price   <- params[4]

  utilities <- with(data,
    beta_netflix * netflix +
    beta_prime   * prime +
    beta_ads     * ad +
    beta_price   * price
  )

  # Exponentiated utilities per task
  data$exp_util <- exp(utilities)

  # Sum of exp utilities per choice set (chid)
  data <- data %>%
    group_by(chid) %>%
    mutate(denom = sum(exp_util)) %>%
    ungroup()

  # Probability of choosing each alternative
  data$prob <- data$exp_util / data$denom

  # Log-likelihood: sum only over chosen alternatives
  log_lik <- sum(log(data$prob[data$choice == 1]))

  # Return negative log-likelihood for minimization
  return(-log_lik)
}

# _todo: Use `optim()` in R or `scipy.optimize()` in Python to find the MLEs for the 4 parameters ($\beta_\text{netflix}$, $\beta_\text{prime}$, $\beta_\text{ads}$, $\beta_\text{price}$), as well as their standard errors (from the Hessian). For each parameter construct a 95% confidence interval._

# Starting values
start_vals <- c(0, 0, 0, 0)

# Run optimization
fit <- optim(
  par = start_vals,
  fn = log_likelihood,
  data = X_data,
  hessian = TRUE,
  method = "BFGS"
)

# MLE estimates
mle_estimates <- fit$par
names(mle_estimates) <- c("β_netflix", "β_prime", "β_ads", "β_price")

# Step 4

# Variance-covariance matrix from inverse Hessian
vcov_matrix <- solve(fit$hessian)
std_errors <- sqrt(diag(vcov_matrix))

# 95% Confidence Intervals
z <- qnorm(0.975)
ci_lower <- mle_estimates - z * std_errors
ci_upper <- mle_estimates + z * std_errors

# Combine into a table
results <- data.frame(
  Estimate = mle_estimates,
  Std.Error = std_errors,
  CI.Lower = ci_lower,
  CI.Upper = ci_upper
)


# Step 5

library(knitr)
kable(results, digits = 3, caption = "MLE Estimates with 95% Confidence Intervals")
```


```{r, echo=FALSE, message=FALSE, warning=FALSE}

# Step 6A

library(ggplot2)

results <- results %>%
  mutate(Parameter = rownames(results))

ggplot(results, aes(x = Parameter, y = Estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = CI.Lower, ymax = CI.Upper), width = 0.2) +
  labs(title = "MLE Coefficient Estimates with 95% Confidence Intervals",
       y = "Estimate", x = "Parameter") +
  theme_minimal()
```

Netflix has the strongest positive effect on choice, followed by Prime. Ads reduce the likelihood of choosing a plan, and higher prices are associated with lower utility. All of these effects are statistically significant and align with what we might intuitively expect based on consumer behavior.

To understand how the model's estimates translate into actual consumer choices, we simulate predicted market shares for three streaming plans using the MLE coefficients. Each option combines a brand, ad experience, and price.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

# Step 6B

predict_share <- function(profiles, coef) {
  utilities <- with(profiles,
    coef["β_netflix"] * netflix +
    coef["β_prime"]   * prime +
    coef["β_ads"]     * ad +
    coef["β_price"]   * price
  )
  exp_util <- exp(utilities)
  probs <- exp_util / sum(exp_util)
 
}

# Step 6C

# Simulate 3 competing options
profiles <- data.frame(
  netflix = c(1, 0, 0),
  prime   = c(0, 1, 0),
  ad      = c(0, 1, 0),
  price   = c(12, 8, 10)
)

# Predict market shares
predict_share(profiles, mle_estimates)

# Step 6D

sim_shares <- predict_share(profiles, mle_estimates)

bar_df <- data.frame(
  Alternative = c("Netflix, No Ads, $12", "Prime, Ads, $8", "Hulu, No Ads, $10"),
  Share = sim_shares
)

ggplot(bar_df, aes(x = Alternative, y = Share)) +
  geom_col(fill = "steelblue") +
  labs(title = "Simulated Choice Shares Based on MLE Estimates",
       y = "Predicted Share", x = "") +
  theme_minimal()

```

In this simulation, Netflix has the highest predicted share, even though it’s the most expensive. This suggests people value the brand and ad-free experience more than just a low price. Prime has the lowest share, likely because it includes ads.


## 5. Estimation via Bayesian Methods

We next estimate the same parameters using Bayesian inference via a Metropolis-Hastings MCMC sampler. The table below shows posterior means, standard deviations, and 95% credible intervals across 10,000 retained samples.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

# _todo: code up a metropolis-hasting MCMC sampler of the posterior distribution. Take 11,000 steps and throw away the first 1,000, retaining the subsequent 10,000._

# _hint: Use N(0,5) priors for the betas on the binary variables, and a N(0,1) prior for the price beta._


log_prior <- function(params) {
  dnorm(params[1], 0, sqrt(5), log = TRUE) +  # beta_netflix
  dnorm(params[2], 0, sqrt(5), log = TRUE) +  # beta_prime
  dnorm(params[3], 0, sqrt(5), log = TRUE) +  # beta_ads
  dnorm(params[4], 0, 1, log = TRUE)          # beta_price
}

# _hint: instead of calculating post=lik*prior, you can work in the log-space and calculate log-post = log-lik + log-prior (this should enable you to re-use your log-likelihood function from the MLE section just above)_


log_posterior <- function(params, data) {
  -log_likelihood(params, data) + log_prior(params)
}

# _hint: King Markov (in the video) use a candidate distribution of a coin flip to decide whether to move left or right among his islands.  Unlike King Markov, we have 4 dimensions (because we have 4 betas) and our dimensions are continuous.  So, use a multivariate normal distribution to pospose the next location for the algorithm to move to. I recommend a MNV(mu, Sigma) where mu=c(0,0,0,0) and sigma has diagonal values c(0.05, 0.05, 0.05, 0.005) and zeros on the off-diagonal.  Since this MVN has no covariances, you can sample each dimension independently (so 4 univariate normals instead of 1 multivariate normal), where the first 3 univariate normals are N(0,0.05) and the last one if N(0,0.005)._


set.seed(123)
n_iter <- 11000
params <- matrix(NA, nrow = n_iter, ncol = 4)
params[1, ] <- c(0, 0, 0, 0)

proposal_sd <- c(0.05, 0.05, 0.05, 0.005)

accept <- rep(NA, n_iter)

for (i in 2:n_iter) {
  current <- params[i - 1, ]
  proposal <- rnorm(4, mean = current, sd = proposal_sd)

  log_post_current <- log_posterior(current, X_data)
  log_post_proposal <- log_posterior(proposal, X_data)

  log_accept_ratio <- log_post_proposal - log_post_current

  if (log(runif(1)) < log_accept_ratio) {
    params[i, ] <- proposal
    accept[i] <- 1
  } else {
    params[i, ] <- current
    accept[i] <- 0
  }
}
```

The trace plot helps us check whether our Bayesian model worked properly. Each dot shows a sampled value for the price coefficient at a given step. If the model is working well, the values should bounce around randomly and not drift up or down over time.


```{r, echo=FALSE, message=FALSE, warning=FALSE}

# todo: for at least one of the 4 parameters, show the trace plot of the algorithm

burn_in <- 1000
post_samples <- params[(burn_in + 1):n_iter, ]
colnames(post_samples) <- c("beta_netflix", "beta_prime", "beta_ads", "beta_price")

# Trace plot for beta_price
library(ggplot2)

df_price <- data.frame(iter = 1:nrow(post_samples), value = post_samples[, "beta_price"])

ggplot(df_price, aes(x = iter, y = value)) +
  geom_line() +
  labs(title = "Trace Plot: beta_price", y = "Value", x = "Iteration") +
  theme_minimal()
```

The trace plot looks relatively healthy. The values move up and down around a central number without any strange trends. This means the sampling process worked well and the results are trustworthy.

This histogram shows the range of values our model thinks the price coefficient could be. It reflects our uncertainty about that number, but if most values are clustered in one spot, it means we’re pretty confident.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

# todo: for at least one of the 4 parameters, show the histogram of the posterior distribution.

# Histogram for beta_price

ggplot(df_price, aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  labs(title = "Posterior Distribution: beta_price", x = "Value", y = "Frequency") +
  theme_minimal()
```

Most of the values are tightly packed around -0.10, which confirms that we’re confident that higher prices lower the likelihood of someone choosing a plan.

We then used Bayesian modeling to estimate these same preferences. This method doesn’t just give one number, but instead gives a full distribution of likely values. The table shows the average, uncertainty, and intervals for each coefficient.


```{r, echo=FALSE, message=FALSE, warning=FALSE}

# _todo: report the 4 posterior means, standard deviations, and 95% credible intervals

posterior_summary <- apply(post_samples, 2, function(x) {
  c(mean = mean(x), sd = sd(x), 
    CI_lower = quantile(x, 0.025), 
    CI_upper = quantile(x, 0.975))
})

posterior_summary <- t(round(posterior_summary, 3))
knitr::kable(posterior_summary, caption = "Posterior Summary (Bayesian MCMC)")
```

The Bayesian results closely match the earlier MLE estimates. Price, for example, still has a strong and very consistent negative effect. The similarity between the two methods builds confidence in our findings.

To make it easy to compare both methods side-by-side, the table below shows how each coefficient lined up between MLE and Bayesian estimation.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

# _todo: Compare them to your results from the Maximum Likelihood approach

# Create side-by-side format
compare_df <- data.frame(
  Estimate = colnames(post_samples),
  MLE = round(mle_estimates, 3),
  Bayesian = round(colMeans(post_samples), 3)
)

# Print table
knitr::kable(compare_df, caption = "Comparison of MLE and Bayesian Estimates", align = "lcc")
```

As shown, the numbers are nearly identical across both approaches. This gives us extra assurance that the patterns we see in the data are real and not a fluke of the estimation method.


## 6. Discussion

```{r, echo=FALSE, message=FALSE, warning=FALSE}

# _todo: Suppose you did not simulate the data. What do you observe about the parameter estimates? What does $\beta_\text{Netflix} > \beta_\text{Prime}$ mean? Does it make sense that $\beta_\text{price}$ is negative?_

# _todo: At a high level, discuss what change you would need to make in order to simulate data from --- and estimate the parameters of --- a multi-level (aka random-parameter or hierarchical) model. This is the model we use to analyze "real world" conjoint data._

```


### **What do the parameter estimates tell us?**

If we didn’t know this data was simulated, the parameter estimates would suggest clear patterns in people’s preferences for streaming services.

- The estimate for $\beta_\text{Netflix}$ is higher than $\beta_\text{Prime}$, which means respondents generally preferred Netflix the most, followed by Prime. Since Hulu was treated as the baseline (reference group), these values tell us that both Netflix and Prime are more attractive than Hulu.

- The estimate for $\beta_\text{price}$ is negative, meaning that as the price of a streaming plan goes up, the likelihood of someone choosing it goes down. This result reflects basic price sensitivity and aligns with how people typically make decisions in markets like streaming subscriptions.

- The negative coefficient for ads shows that including ads makes a plan less appealing, which makes intuitive sense since many consumers value an ad-free experience.

Overall, the estimates appear reasonable and consistent with how people are likely to evaluate trade-offs among brand, price, and ad experience.

### **How would this work in a real-world setting?**

In the real world, people don’t all value things the same way. Some might care more about price, others about brand or whether the plan includes ads. The current model assumes that everyone has the same preferences, but in practice, that’s not always true.

To reflect this variation, we would need to use a multi-level model/hierarchical model. This type of model allows each individual to have their own set of preference weights ($\beta$s), while still assuming those individual preferences come from a broader population distribution.

To simulate and estimate a multi-level model, we would:

- **Simulation**: Start by drawing a unique $\beta$ vector for each respondent from a population-level distribution (like a multivariate normal). Then simulate choices for each respondent using their own $\beta$s.

- **Estimation**: Use Bayesian methods (like MCMC) to estimate both the distribution of individual-level preferences and the overall population-level averages. 

This approach is especially useful when analyzing real-world conjoint data, where understanding differences across people is just as important as understanding overall trends. It enables better targeting, segmentation, and personalized product design.

